\documentclass[12pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage[margin=1in]{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
%\usepackage{cite}
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{url}

\title{Predicting Chemical Fate Using Deep Learning}
\author{Zach Calhoun}
\date{March 4, 2022}	

\begin{document}
\maketitle

%TODO - match formatting laid out in project description
\section{Summary}
Machine learning is rapidly gaining traction as a powerful tool to expedite the drug discovery process \cite{Jiang:2021}. However, its adoption within ecotoxicology is lagging behind \cite{miller:2018}.

\section{Problem Statement}
Roughly 10 million new chemicals are synthesized each year, and the EPA is only equipped to screen a fraction of these chemicals for environmental risk \cite{burton}. Of particular concern are so-called PBT (persistent, bioaccumulative, and toxic) chemicals. These chemicals are unique because they do not break down, they are particularly toxic, and because they bioaccumulate, they risk making their way into the food chain \cite{matthies}.

Perhaps the most infamous example of a PBT is DDT, the fertilizer used during the 20th century that was found to be highly toxic to bird populations, inspiring Rachel Carson's \emph{Silent Spring}. The more recent example making headlines today are PFAS (per-fluorinated alkyl substances), which are chemicals commonly found in non-stick pans or in water-repellant clothing. In both cases, the chemicals were synthesized for a particular purpose, and the risk to society was discovered decades later. Decades later, however, is far too late. Both PFAS and DDT are now ubiquitous, found in remote locations such as the Arctic, as well as in a large percentage of the population's bloodstream \cite{czub}.

To prevent this pattern from repeating itself, the EPA conducts environmental risk assessments using known and estimated chemical properties to predict chemical fate and toxicity in the environment. However, the EPA can only fully study a fraction of the compounds they need to, and estimations can be unreliable \cite{burton}. For example, the Bond Contribution Method is frequently used to predict Henry's law \cite{meylan}. This method is based on a small set of chemicals, and its output may be unreliable, since it only focuses on the bonds in the molecule rather than the structure as a whole. Deep learning, and specifically, Graph Neural Networks and Variational Auto-encoders both represent models that may be able to more effectively represent the impact of molecular structure on Henry's law.

If we could better predict this property, then environmental chemists could more accurately assess the true risk of new compounds on the market, allowing society to better prevent the widespread adoption of chemicals that pose significant environmental threats.

\section{Data Summary}
Henry's law (also called the air/water partitioning coefficient) is chosen because it is a well studied partitioning coefficient, meaning a lot of chemical's have values based on observations rather than estimations. I will reference two data sources to get started.
\begin{enumerate}
\item EPA CompTox Dashboard. The EPA maintains a database of over 350,000 chemicals and their chemical properties. From this dashboard, we would like to restrict our analysis to only the chemicals that are organic and that contain an observed value for Henry's law (rather than an estimated value). The organic compound restriction is placed to reduce the complexity of the graphs used in our analysis \cite{epa-comptox}. 
\item Meylan et al's 1991 paper on the Bond Contribution Method for estimating Henry's law \cite{meylan}. This paper includes data for 345 organic compounds, on which the coefficients for the Bond Contribution Method are based. The Bond Contribution method is still referenced in the literature today as an estimation method for Henry's law, so performance will be benchmarked against this method. 

.% Is this still referenced in the literature?
\end{enumerate}

\section{Approach}
Machine Learning promises to improve the speed of chemical discovery and understanding. However, there is still much debate as to the best method for analyzing chemicals \cite{Jiang:2021}. Specifically, descriptor based representations of molecules (i.e., data about the molecule like its molecular weight and number of atoms), have been shown to outperform more complicated graph neural network representations of these molecules.

However, few studies have explored which approach better models properties that are important for ecotoxicology. In this report, we want to explore whether we can model a common physico-chemical property well using Graph Neural Nets. 

\section{Discussion}

\subsection{Suggestions for future research}
As explored in \cite{Stojic:2019}, machine learning has been applied to explore wet deposition of specific compounds. The theoretical Henry's law constant can explain a large percentage of the chemical partitioning behavior between air and water. However, in practice, partitioning behavior depends on a wide variety of factors: for example, pH or the presence of salt can change partitioning behavior drastically.

The purpose of this project is to study a framework for modeling this complex behavior, but its scope was limited to predicting the observed Henry's Law constant at standard temperature and pressure. Thus, we propose that further studies explore how environmental factors can be incorporated into this model to generalize beyond standard assumptions. Further development would result in a method that could be applied to real-world problems.


\bibliography{ds4cee-final-project}{}
\bibliographystyle{ieeetr}

\end{document}  